Is high social class always beneficial for survival? Northern Sweden 1801--2013

### 2018-02-06 article3b.Rmd
fix out.height

---
title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
author: Sören Edvinsson\footnote{Corresponding author; Phone number +46907866040, email soren.edvinsson\@umu.se} and Göran Broström
date: 'Centre for Demographic and Ageing Research \newline Umeå University, SE-90187 Umeå, Sweden'
output:
  bookdown::pdf_document2:
    citation_package: natbib
    number_sections: yes
    toc: no
    toc_depth: 2
    keep_tex: true
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
titlepage: yes
bibliography: "mybib.bib"
documentclass: article
classoption: titlepage
fontsize: 12pt
citation_package: natbib
biblio-style: "apalike"
header-includes:
  - \usepackage[utf8]{inputenc}
abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common notions of the assumed advantage in survival of belonging to a high social class. The issue is analysed according to gender and age group (adults and elderly) and in relation to the development of economic inequality. The results show that high social class is not always favourable for survival. Men in the elite category had higher mortality compared to others during a large part of the studied period; a male mortality class reversal appears at a surprisingly late date, while the social gradient among women conforms to the expected pattern. We suggest that health-related behaviour is decisive not only in later but earlier phases of the mortality transition as well. The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology."
---

@@

```{r mapp,fig.cap = "(ref:cities)", fig.height=6,fig.align='center', out.height='50%'}
knitr::include_graphics("myfigs/sweden.png")
```
During the studied period, Sweden developed from a poor agricultural society with low urbanization to a rich welfare state. If Sweden is remote from a European continental perspective, the regions studied here are remote from the central parts of the country. The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the county of Västerbotten in northern of Sweden along the coast of the Gulf
of Bothnia. This is a remote part of the country where communication with
the rest of Sweden was difficult until the late 19th century. The economy
was dominated by agriculture, making it vulnerable to harvest failures; several severe famines occurred in the regions during the 1800s, for example after the harvest failure of 1867 [@segb14]. During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen. The first ships of the year came to this area in May or even as late as June in some cases [@fah56]. Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, facilitating contact with the rest of Sweden and, improving the economy and making it possible to mitigate the effects of harvest failures. During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden.

@@

```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, missing) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = '50%', fig.height=4}
xx <- with(skel[skel$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xy <- with(skume[skume$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xz <- with(foli[foli$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xxx, 1) * 100, 1)
barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
```

```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, women.", out.height = '50%', fig.scap="Exposure by social class and time period, women.",fig.height=4}
xx <- with(skel[skel$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xz <- with(foli[foli$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xxx, 1) * 100, 1)
barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
```

@@

```{r civstskum, fig.cap = "Exposure by marital status (from the top: never married, married and previously married) and time period", fig.scap='Exposure by marital status and time period.',out.height = '50%',fig.height=4}
xx <- with(skel, tapply(exposure, list(period = period, civst = civst), sum))
xy <- with(skume, tapply(exposure, list(period = period, civst = civst), sum))
xz <- with(foli, tapply(exposure, list(period = period, civst = civst), sum))
xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xxx, 1) * 100, 1)
barplot(t(yy)[3:1, ], col = c("red", "blue", "green"), las = 2, ylab = "Per cent")
```

>>>>

---
title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
author: Sören Edvinsson\footnote{Corresponding author; Phone number +46907866040, email soren.edvinsson\@umu.se} and Göran Broström
date: 'Centre for Demographic and Ageing Research \newline Umeå University, SE-90187 Umeå, Sweden'
output:
  bookdown::pdf_document2:
    citation_package: natbib
    number_sections: yes
    toc: no
    toc_depth: 2
    keep_tex: true
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
titlepage: yes
bibliography: "mybib.bib"
documentclass: article
classoption: titlepage
fontsize: 12pt
citation_package: natbib
biblio-style: "apalike"
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{graphicx}
abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common notions of the assumed advantage in survival of belonging to a high social class. The issue is analysed according to gender and age group (adults and elderly) and in relation to the development of economic inequality. The results show that high social class is not always favourable for survival. Men in the elite category had higher mortality compared to others during a large part of the studied period; a male mortality class reversal appears at a surprisingly late date, while the social gradient among women conforms to the expected pattern. We suggest that health-related behaviour is decisive not only in later but earlier phases of the mortality transition as well. The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology."
---

@@

```{r mapp,fig.cap = "(ref:cities)", fig.height=6,fig.align='center', out.height="5in"}
knitr::include_graphics("myfigs/sweden.png")
```
During the studied period, Sweden developed from a poor agricultural society with low urbanization to a rich welfare state. If Sweden is remote from a European continental perspective, the regions studied here are remote from the central parts of the country. The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the county of Västerbotten in northern of Sweden along the coast of the Gulf
of Bothnia. This is a remote part of the country where communication with
the rest of Sweden was difficult until the late 19th century. The economy
was dominated by agriculture, making it vulnerable to harvest failures; several severe famines occurred in the regions during the 1800s, for example after the harvest failure of 1867 [@segb14]. During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen. The first ships of the year came to this area in May or even as late as June in some cases [@fah56]. Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, facilitating contact with the rest of Sweden and, improving the economy and making it possible to mitigate the effects of harvest failures. During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden.

@@

```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, missing) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = "14cm", fig.height=4}
xx <- with(skel[skel$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xy <- with(skume[skume$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xz <- with(foli[foli$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xxx, 1) * 100, 1)
barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
```

```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4}
xx <- with(skel[skel$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xz <- with(foli[foli$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xxx, 1) * 100, 1)
barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
```

@@

```{r civstskum, fig.cap = "Exposure by marital status (from the top: never married, married and previously married) and time period", fig.scap='Exposure by marital status and time period.',out.height = "14cm",fig.height=4}
xx <- with(skel, tapply(exposure, list(period = period, civst = civst), sum))
xy <- with(skume, tapply(exposure, list(period = period, civst = civst), sum))
xz <- with(foli, tapply(exposure, list(period = period, civst = civst), sum))
xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xxx, 1) * 100, 1)
barplot(t(yy)[3:1, ], col = c("red", "blue", "green"), las = 2, ylab = "Per cent")
```
